---
title: "Climate change and temperature anomalies"
author: "Ishaan Khetan, Kathlyn Lee, Emilia Moskala, Juan Sanchez-Blanco, Sylvie Zheng, Jingyi Fang"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(wbstats)
library(countrycode)
library(patchwork)
library(gganimate)
library(infer)
```

```{r weather_data, cache=TRUE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v4/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")

```

```{r tidyweather}
# SELECT YEAR, MONTHS AND PIVOT LONGER
weather %>% 
  select(1:13) %>% 
  pivot_longer(2:13, names_to = "month", values_to = "delta") -> tidyweather
```

## Plotting Information

```{r scatter_plot}

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), month, "1")),
         month = month(date, label=TRUE),
         Year = year(date))

ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "Weather Anomalies",
    x = "Year",
    y = "Delta"
  )

```

```{r facet_wrap}

ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  facet_wrap(~month) +
  theme_bw() +
  labs (
    title = "Weather Anomalies",
    x = "Year",
    y = "Delta"
  )

```

> There is no apparent difference between the effects of increasing
> temperature in monthly data. January and February seem to represent
> the strongest change, but it does not vary significantly from other
> months.

```{r intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))
```

```{r density_plot}

comparison %>% 
  ggplot(aes(x = delta, fill = interval, color = interval)) +
  geom_density(alpha = 0.3) +
  labs (
    title = "Weather Delta by Decade",
    x = "Delta",
    y = "Density",
    fill = "Decade",
    color = "Decade"
  )
```

```{r averaging}

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   #grouping data by Year
  # creating summaries for mean delta 
  # use `na.rm=TRUE` to eliminate NA (not available) values 
  summarise(mean_delta = mean(delta, na.rm=TRUE)) 

average_annual_anomaly %>% 
  ggplot(aes(x=Year, y=mean_delta)) +
  geom_point(aes(color=mean_delta>0)) + # DRAWING POINTS ABOVE ZERO A DIFFERENT COLOUR
  geom_smooth(method = "loess", color="black") +
  theme_bw() + 
  labs(
    title = "Weather Delta average by Year",
    x = "Year",
    y = "Mean Delta", 
    color = "Above Zero"
  )
```

## Confidence Interval for `delta`

#### CI using formula

```{r, calculate_CI_using_formula}

formula_ci <- comparison %>% 

  # choose the interval 2011-present
  filter(Year >= 2011) %>% 
  group_by(Year) %>% 
  summarise(
    mean_delta = mean(delta),
    sd_delta = sd(delta),
    count = n(),
    # We're choosing a 95% confidence interval:
    t_critical = qt(0.975, count-1),
    se_delta = sd(delta/sqrt(count)),
    margin_of_error = t_critical*se_delta,
    delta_low  = mean_delta - margin_of_error,
    delta_high = mean_delta + margin_of_error
  )
  
  # calculate summary statistics for temperature deviation (delta) 
  # calculate mean, SD, count, SE, lower/upper 95% CI
  # what dplyr verb will you use? 

#print out formula_CI
formula_ci
```

> In order to calculate the confidence intervals, we have calculated for
> every year many summary statistics including mean, SD, SE and sample
> size. Using these summary statistics we have calculated the t-Student
> distribution critical value, and multiplied the `t_critical` by the
> standard error to get the final margin of error. Finally we add and
> subtract the margin of error to the mean to get the limits of the
> interval. In order to fully understand the data we have collected, we
> have come up with the following visualization of the mean and
> confidence intervals:

```{r}
formula_ci %>% 
  mutate(Year = as.factor(Year)) %>% 
  na.omit() %>% 
  ggplot(aes(color = Year)) +
  geom_pointrange(aes(x=Year, y=mean_delta, ymax=delta_high, ymin=delta_low)) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(
    title = "Weather Delta range by Year",
    subtitle = "NASA Weater Data",
    x = "Year",
    y = "Delta range", 
    color = NULL
  )
```

> The visualization shows an increase in the mean over the years,
> reaching a first maximum in 2016, and an even higher maximum mean in
> 2020. For years with a higher mean delta, the confidence intervals
> seem to increase as well. With the current samples and intervals, we
> cannot say for sure that the mean delta in year 2016 was higher than
> the mean delta in 2017 for instance, but we can be sure with 95%
> confidence that 2011 was lower than 2016.

#### CI using formula (for all years)

Since there are not enough data points every year, the confidence
intervals are very large and vary a lot by year. The total confidence
interval for the entire period is as follows:

```{r}
formula_ci_interval <- comparison %>% 

  # choose the interval 2011-present
  filter(Year >= 2011) %>% 
  na.omit() %>% 
  group_by(interval) %>% 
  summarise(
    mean_delta = mean(delta),
    sd_delta = sd(delta),
    count = n(),
    # We're choosing a 95% confidence interval:
    t_critical = qt(0.975, count-1),
    se_delta = sd(delta/sqrt(count)),
    margin_of_error = t_critical*se_delta,
    delta_low  = mean_delta - margin_of_error,
    delta_high = mean_delta + margin_of_error
  )

#print out formula_CI
formula_ci_interval %>% select(delta_low, delta_high)
```

#### CI using bootstrapping

```{r, calculate_CI_using_bootstrap}

boot_dist <- comparison %>%
  # choose the interval 2011-present
  filter(Year >= 2011) %>%
  mutate(Year = as.factor(Year)) %>%
  specify(response=delta) %>%
  generate(reps=1000, type="bootstrap") %>% 
  calculate(stat = "mean")

boot_dist %>%
  # Calculate the confidence interval around the point estimate
  get_confidence_interval(
    # At the 95% confidence level; percentile method
    level = 0.95
  )
```
